---
title: "Knapsack Package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Knapsack Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
  #eval = FALSE
)

options(expressions = 50000)
```
<br>

## Package description
The package "knapsack" is based on laboratory 6 of the course "Advanced R programming" at LiU. It implements 3 different algorithms to solve or at least approximate the optimal solution for the famous knapsack optimization problem: 

|   Algorithm    |  Description  |   Complexity   |  
|:---------------|:--------------|---------------:|
|  Brute Force       |  Tries out all $2^n$ possible item-combinations  |  $O(2^n)$    |  
|  Dynamic Programming   |  Uses dynamic programming for solving the problem  |  $O(Wn)$      |  
|  Greedy Heuristic    |  Heuristic approach to find a good approximation of the optimal solution |  $O(n\ log(n))$     |  

In the following paragraphs, we demonstrate the proper usage of the packages functions and also analyze the time complexity of the different algorithms.

<br>

## Brute Force Algorithm

The following code shows two examples of the brute force algorithm.

```{r brute}
library(knapsack)


# we create a dataframe with 2000 possible knapsack items 
knapsack_objects <- get_knapsack_objects(2000)


# Example 1:

# now the brute force algorithm is used to solve for the first 8 items with a 
# maximum weight W of 3500

brute_force_knapsack(x = knapsack_objects[1:8, ], W = 3500)


# Example 2:

# now the brute force algorithm is used to solve for the first 12 items with a 
# maximum weight W of 2000

brute_force_knapsack(x = knapsack_objects[1:12,], W = 2000)

```

### Computation time - Brute Force
```{r}
# number of items = 16, W = 3500
time <- system.time(brute_force_knapsack(x = knapsack_objects[1:16, ], W = 3500))
time
```

```{r, echo=FALSE}
cat(sprintf(" Computation time for Brute Force (16 items): %#.4f seconds\n\n", time[3]), 
    sprintf("Computation time for Brute Force (18 items): %#.4f seconds\n\n",  system.time(brute_force_knapsack(x = knapsack_objects[1:18, ], W = 3500))[3]),
    sprintf("Computation time for Brute Force (20 items): %#.4f seconds\n\n", system.time(brute_force_knapsack(x = knapsack_objects[1:20, ], W = 3500))[3]))
```

<br>

## Dynamic Programming

The following code shows the same two examples as before but uses the dynamic programming algorithm.

```{r dynamic}
# Example 1:

# now the dynamic programming algorithm is used to solve for the first 8 items with a 
# maximum weight W of 3500

knapsack_dynamic(x = knapsack_objects[1:8, ], W = 3500)


# Example 2:

# now the dynamic programming algorithm is used to solve for the first 12 items with a 
# maximum weight W of 2000

knapsack_dynamic(x = knapsack_objects[1:12,], W = 2000)

```

### Computation time - Dynamic Programming
```{r}
# number of items = 125, W = 3500
time <- system.time(knapsack_dynamic(x = knapsack_objects[1:125, ], W = 3500))
time
```

```{r, echo=FALSE}
cat(sprintf(" Computation time for Dynamic Programming (125 items): %#.4f seconds\n\n", time[3]), 
    sprintf("Computation time for Dynamic Programming (250 items): %#.4f seconds\n\n",  system.time(knapsack_dynamic(x = knapsack_objects[1:250, ], W = 3500))[3]),
    sprintf("Computation time for Dynamic Programming (500 items): %#.4f seconds\n\n", system.time(knapsack_dynamic(x = knapsack_objects[1:500, ], W = 3500))[3]))
```

<br>

## Greedy Heuristic

The following code shows two examples for the usage of the greedy heuristic, with a higher amount of items. 

```{r greedy}
# Example 1:

# now the greedy heuristic is used to solve for the first 800 items with a 
# maximum weight W of 3500

greedy_knapsack(x = knapsack_objects[1:800,], W = 3500)


# Example 2:

# now the greedy heuristic is used to solve for the first 1200 items with a 
# maximum weight W of 2000

greedy_knapsack(x = knapsack_objects[1:1200,], W = 2000)

```

#### Computation time - Greedy Heuristic
```{r}
# we create a dataframe with 1,000,000 possible knapsack items 
knapsack_objects <- get_knapsack_objects(1000000)

# number of items = 10000, W = 3500
time <- system.time(greedy_knapsack(x = knapsack_objects[1:10000, ], W = 3500))
print(time)
```

```{r, echo=FALSE}
cat(sprintf(" Computation time for Greedy Heuristic (10,000 items): %#.4f seconds\n\n", time[3]), 
    sprintf("Computation time for Greedy Heuristic (100,000 items): %#.4f seconds\n\n",  system.time(greedy_knapsack(x = knapsack_objects[1:100000, ], W = 3500))[3]),
    sprintf("Computation time for Greedy Heuristic (1,000,000 items): %#.4f seconds\n\n", system.time(greedy_knapsack(x = knapsack_objects, W = 3500))[3]))
```


<br>

## Optimization of Computational Time

After profiling the two algorithms, Brute Force and Dynamic Programming, an optimized version of both of them was created, to reduce bottlenecks and thus computational time. 

### Optimized Brute Force Algorithm

Profiling showed, that transposig the "combi-vector" takes a lot of time. But it is just a necessary operation in math but not in programming, just multiplying both vectors yields the same result. So we simply deleted this operation, which lead to reduced compuitational time:

```{r}
# number of items = 20, W = 3500
time_1 <- system.time(brute_force_knapsack(x = knapsack_objects[1:20, ], W = 3500))
print(time_1)

time_2 <- system.time(optimized_brute_force(x = knapsack_objects[1:20, ], W = 3500))
print(time_2)
```

```{r, echo=FALSE}
cat(sprintf(" Computation time for Normal Brute Force Algorithm (20 items): %#.4f seconds\n\n", time_1[3]), 
    sprintf("Computation time for Optimized Brute Force Algorithm (20 items): %#.4f seconds\n\n",  time_2[3]),
    sprintf("Improvement in computational time: %#.4f seconds\n\n", time_1[3]-time_2[3]))
```


### Optimized Dynamic Programming Algorithm

Profiling showed, that the  operation  "*$*" takes a lot of time. In the for loop we use it twice to refer to the same value in the dataframe, that is why we can simply store it in a variable, to reduce the number of times the operation is used:

```{r}
# number of items = 1500, W = 3500
time_1 <- system.time(knapsack_dynamic(x = knapsack_objects[1:1500, ], W = 3500))
print(time_1)

time_2 <- system.time(optimized_dynamic_programming(x = knapsack_objects[1:1500, ], W = 3500))
print(time_2)
```

```{r, echo=FALSE}
cat(sprintf(" Computation time for Normal Dynamic Programming Algorithm (1500 items): %#.4f seconds\n\n", time_1[3]), 
    sprintf("Computation time for Optimized Dynamic Programming Algorithm (1500 items): %#.4f seconds\n\n",  time_2[3]),
    sprintf("Improvement in computational time: %#.4f seconds\n\n", time_1[3]-time_2[3]))
```


### Parallelization of Brute Force Algorithm

To analyze the impact of parallelization on computation time, we integrated an optional parallelized solving approach to the Brute Force Algorithm that uses the *mcapply* function. By default, *brute_force_knapsack* still uses non-parallelized processes, but by setting setting the argument *parallel* to `r TRUE`, the parallelized approach is used.

The following code contains two examples, which show the time efficiency gain due to parallelization:

```{r parallel}
# NORMAL, number of items = 18, W = 3500
system.time(brute_force_knapsack(x = knapsack_objects[1:18, ], W = 3500, parallel = FALSE))


# PARALLEL, number of items = 18, W = 3500
system.time(brute_force_knapsack(x = knapsack_objects[1:18, ], W = 3500, parallel = TRUE))
```

```{r}
# NORMAL, number of items = 23 W = 30000
system.time(brute_force_knapsack(x = knapsack_objects[1:23, ], W = 30000, parallel = FALSE))


# PARALLEL, number of items = 23, W = 30000
system.time(brute_force_knapsack(x = knapsack_objects[1:23, ], W = 30000, parallel = TRUE))
```

